---
title: "AMS 580 Midterm 1"
author: "Kachun Huang"
date: "2025-03-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# AMS 580 Exam 1

## Import Library

```{r}
library(tidyverse)
library(caret)
library(neuralnet)
library(leaps)
library(bestglm)
library(MASS)
library(dplyr)
```

# Task 1

## Import Data

```{r}
unknown <- read.csv("/Users/thomashuang/Downloads/Unknown.csv")
head(unknown)
```

```{r}
str(unknown)
```

## Data Pre-processing

### Remove Missing Values

```{r}
unknown <- na.omit(unknown)
head(unknown)
```

```{r}
str(unknown)
```

After cleaning missing values, there are 4621 observations left.

## Data Splitting

```{r}
set.seed(123)
training.sample <- unknown$y %>% createDataPartition(p=0.75, list=FALSE)
train.data <- unknown[training.sample,]
test.data <- unknown[-training.sample,]
```

## NN Model with 1 hidden

```{r}
nn_one_model_task1 <- neuralnet(
  y ~.,
  train.data,
  hidden = 3,
  err.fct = "ce",
  linear.output = FALSE
)
```

```{r}
plot(nn_one_model_task1, rep = "best")
```

### Predict on Test Data

```{r}
pred_nn_one_task1 <- predict(nn_one_model_task1, test.data)
pred_nn_one_task1_class <- ifelse(pred_nn_one_task1 > 0.5,1,0)
```

### Confusion Matrix

```{r}
cm1 <- confusionMatrix(factor(pred_nn_one_task1_class), factor(test.data$y), positive = "1")
cm1
```

### sensitivity

```{r}
cm1$byClass["Sensitivity"]
```

### specificity

```{r}
cm1$byClass["Specificity"]
```

### overall accuracy

```{r}
cm1$overall["Accuracy"]
```

## Logistic Model

```{r}
log_model <- glm(y ~., data=train.data, family = binomial)
summary(log_model)
```

### Predict on Test Data

```{r}
pred_prob <- log_model %>% predict(test.data, type="response")
predicted.classes <- ifelse(pred_prob > 0.5, "1", "0")
```

### Confusion Matrix

```{r}
cm2 <- confusionMatrix(factor(predicted.classes),factor(test.data$y), positive = "1")
cm2
```

### sensitivity

```{r}
cm2$byClass["Sensitivity"]
```

### specificity

```{r}
cm2$byClass["Specificity"]
```

### overall accuracy

```{r}
cm2$overall["Accuracy"]
```

## Write Dataset File

```{r}
predictions <- data.frame(
  true_obs = test.data$y,
  nn_pred = pred_nn_one_task1_class,
  log_pred = predicted.classes
)
agreement_table <- table(predictions$nn_pred, predictions$log_pred)
agreement_table
```

### Result

The result said there is a very high agreement rate and low disagreement.

# Task 2

## Import Data

```{r}
madness <- read.csv("/Users/thomashuang/Downloads/Madness.csv")
head(madness)
```

```{r}
str(madness)
```

## Data Pre-processing

### Remove Missing Values

```{r}
madness <- na.omit(madness)
```

```{r}
str(madness)
```

After cleaning, there area 1460 observations left.

### Mean and std

```{r}
mean <- mean(madness$y)
print(mean)
sd <- sd(madness$y)
print(sd)
```

### factorize variable

```{r}
categ_col <- ifelse(madness$x15 == "y", 1, 0)
```

### Scale the Data

```{r}
madness <- data.frame(scale(subset(madness, select = -x15)))

madness$x15 <- categ_col

head(madness)
```

## Data Splitting

```{r}
set.seed(123)
training.sample <- madness$y %>% createDataPartition(p=0.75, list=FALSE)
train.data <- madness[training.sample,]
test.data <- madness[-training.sample,]
```

## NN Model with 1 hidden

```{r}
nn_one_model_task2 <- neuralnet(
  y ~.,
  train.data,
  hidden = 2,
  err.fct = "sse",
  linear.output = TRUE,
  stepmax = 1e6
)
```

### plot

```{r}
plot(nn_one_model_task2, rep="best")
```

### Predict on Test Data

```{r}
pred_one_model_task2 <- predict(nn_one_model_task2, test.data)
```

### Descale

```{r}
# compute y(unscaled)
y <- sd * test.data$y + mean

#compute y^hat (unscaled)
yhat <- pred_one_model_task2*sd+mean
```

### Compute Test MSE

```{r}
sum((y-yhat)^2)/length(y)
```

or

```{r}
(MSE.nn.1 = RMSE(y,yhat)^2)
```

## Plot y\^hat and y

```{r}
plot(y, yhat)
```

## Linear Regression

```{r}
# null model
null <- lm(y ~ 1, data=train.data)

#full model
full <-lm(y ~. , data=train.data)
```

### Bidirectional

```{r}
#find the best model in backward stepwise
fit_step_both_bic <- stepAIC(full, k = log(nrow(train.data)), trace = 1,direction="both")
fit_step_both_bic
```

### Predict on Test Data

```{r}
pred_stepwise <- predict(fit_step_both_bic, test.data)
```

### Descale

```{r}
# compute y(unscaled)
y <- sd * test.data$y + mean

#compute y^hat (unscaled)
yhat <- pred_stepwise*sd+mean
```

### Compute Test MSE

```{r}
sum((y-yhat)^2)/length(y)
```

or

```{r}
(MSE.stepwise = RMSE(y,yhat)^2)
```

## Plot y\^hat and y

```{r}
plot(y, yhat)
```

## Task 2 Result

In the Madness dataset, the stepwise linear regression performs better then neural network with 1 hidden layer and 2 neurons. The possible reason behind this could be

-   Overfitting: with 1 hidden layer and 2 neurons could overfit on the training data and perform worse in testing data

-   Linear Relationship: it is possible that the dataset already explain well in linear model, so neural network might introduces some unnecessary relationship to the model
