---
title: "AMS 580 Final Exam"
author: "Kachun Huang"
date: "2025-05-14"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# AMS 580 Final Exam **Seat Number C5**

# Task 1

## Import Library

```{r,warning = FALSE, message = FALSE}
library(dplyr)
library(caret)
library(randomForest)
library(xgboost)
library(Matrix)
library(keras)
library(tensorflow)
library(reticulate)
library(tidyverse)
library(tidyquant)
library(zoo)
library(magrittr)
```

## Import Data

```{r}
data1 <- read.csv("/Users/thomashuang/Downloads/data_1.csv")
head(data1)
```

```{r}
str(data1)
```

## Data Cleaning

### Remove Missing rows

```{r}
data1 <- na.omit(data1)
str(data1)
```

After Remove NAs, there are 1100 observations left.

## Data Splitting

```{r}
set.seed(2025)
train_idx <- createDataPartition(data1$Anxiety.Level..1.10., p = 0.75, list = FALSE)
train_data1 <- data1[train_idx, ]
test_data1 <- data1[-train_idx, ]
```

For Random Forest

```{r}
train_data1_rf <- train_data1
train_data1_rf$Gender <- factor(train_data1$Gender)
train_data1_rf$Occupation <- factor(train_data1$Occupation)
train_data1_rf$Smoking <- factor(train_data1$Smoking)
train_data1_rf$Family.History.of.Anxiety <- factor(train_data1$Family.History.of.Anxiety)
train_data1_rf$Dizziness  <- factor(train_data1$Dizziness)
train_data1_rf$Medication  <- factor(train_data1$Medication)
train_data1_rf$Recent.Major.Life.Event <- factor(train_data1$Recent.Major.Life.Event )

test_data1_rf <- test_data1
test_data1_rf$Gender <- factor(test_data1$Gender)
test_data1_rf$Occupation <- factor(test_data1$Occupation)
test_data1_rf$Smoking <- factor(test_data1$Smoking)
test_data1_rf$Family.History.of.Anxiety <- factor(test_data1$Family.History.of.Anxiety)
test_data1_rf$Dizziness  <- factor(test_data1$Dizziness)
test_data1_rf$Medication  <- factor(test_data1$Medication)
test_data1_rf$Recent.Major.Life.Event <- factor(test_data1$Recent.Major.Life.Event )
```

For XgBoost

```{r}
# Prepare data for XGBoost
X_train <- model.matrix(Anxiety.Level..1.10. ~., data = train_data1)
y_train <- train_data1$Anxiety.Level..1.10.
X_test <- model.matrix(Anxiety.Level..1.10. ~., data = test_data1)
y_test <- test_data1$Anxiety.Level..1.10.
```

## RandomForest

```{r}
set.seed(2025)

# 10-fold CV
control <- trainControl(
  method = "cv", 
  number = 10
  )

# train the rf using 10-fold CV
model <- caret::train(
  Anxiety.Level..1.10. ~ ., 
  data = train_data1_rf, 
  method = "rf", 
  trControl = control
  )
```

BestTune

```{r}
model$bestTune
```

Variable Importance as barplot

```{r}
final_model <- model$finalModel
barplot(randomForest::importance(final_model)[,"IncNodePurity"])
```

```{r}
set.seed(2025)
rf <- randomForest(
  Anxiety.Level..1.10.~.,
  data=train_data1_rf,
  mtry=16,
  keep.forest=FALSE, #not save each individual CART tree in the output
  importance=TRUE
)
```

### RMSE

```{r}
sqrt(rf$mse[500])
```

## XgBoost

```{r}
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest <- xgb.DMatrix(data = X_test, label = y_test)
```

```{r}
# Train XGBoost
xgb_model <- xgboost(
  data = dtrain,
  nrounds = 500,
  eta = 0.1,
  verbose = 0
)
```

```{r}
# Feature importance plot
importance <- xgb.importance(feature_names = colnames(X_train), model = xgb_model)
xgb.plot.importance(importance, main = "XGBoost Feature Importance")
```

Prediction

```{r}
xgb_preds <- predict(xgb_model, X_test)
```

RMSE

```{r}
sqrt(mean((y_test - xgb_preds)^2))
```

## Comparison

Model Test MSE

-   RF: 1.104186

-   XGB: 1.165522

The reason why RF outperform XGB

-   Number of Features- since XGB used one-hot-encoding so there is a risk of multicolinearity problem which overfit the training data

-   Cross-Validation: since we used cross validation to find the optimal mtry for rf and in XGB, we didn't use CV to tune hyper-parameter. It can result in RF performs better than XGB

# Task 2

## Import Data

```{r}
data2 <- read.csv("/Users/thomashuang/Downloads/data_2.csv")
data2$Date = as.Date(data2$Date)
head(data2)
```

```{r}
str(data2)
```

## 1. Data preprocessing and visualization

```{r}
ggplot(
  data2, 
  aes(x=Date, y = AMZN.Close)
) + geom_line()
```

normalization

```{r}
data2$min_lagged = lag(data2$AMZN.Low)
data2$max_lagged = lag(data2$AMZN.High)

data2$Close_norm = (data2$AMZN.Close - data2$min_lagged) / (data2$max_lagged -data2$min_lagged)
```

```{r}
model_data <- matrix(
  data2$Close_norm[-1] #excluding volume
)
tail(model_data,10)
```

## 2. Data Splitting

```{r}
set.seed(2025)

# Create sliding window dataset
create_dataset <- function(series, timesteps = 10) {
  x <- list()
  y <- c()
  for (i in 1:(length(series) - timesteps)) {
  x[[i]] <- series[i:(i + timesteps - 1)]
  y <- c(y, series[i + timesteps])
  }
  x_array <- do.call(rbind, x)
  return(list(x = x_array, y = y))
}

# Prepare training data
timesteps <- 10

train_data2 <- model_data[1:(length(model_data)-10)] # training = all except last 10 days

test_data2 <- model_data[(length(model_data)-9):length(model_data)] # true last 10 points

train_set2 <- create_dataset(train_data2, timesteps)

x_train_array <- array(
  as.numeric(train_set2$x), 
  dim = c(nrow(train_set2$x), timesteps, 1)
)

y_train_array <- as.numeric(train_set2$y)

x_train2 <- tf$convert_to_tensor(
  x_train_array, 
  dtype = "float32"
)

y_train2 <- tf$convert_to_tensor(
  y_train_array, 
  dtype = "float32"
)

dim(x_train2)
```

The dimension of the training data is 569(rows) and 10(columns)

## LSTM

```{r}
input_lstm <- layer_input(shape = c(timesteps, 1))

output_lstm <- input_lstm %>%
  layer_lstm(units = 32) %>%
  layer_dense(units = 1)

model_lstm <- keras_model(inputs = input_lstm, outputs = output_lstm)

model_lstm$compile(
  loss = "mean_squared_error"
)

history_lstm <- model_lstm$fit(
  x = x_train2,
  y = y_train2,
  epochs = r_to_py(100L),
  batch_size = r_to_py(16L),
  verbose = 0
)
```

## RNN

```{r}
input_rnn <- layer_input(shape = c(timesteps, 1))

output_rnn <- input_rnn %>%
  layer_simple_rnn(units = 32) %>%
  layer_dense(units = 1)

model_rnn <- keras_model(inputs = input_rnn, outputs = output_rnn)

model_rnn$compile(
  loss = "mean_squared_error"
)

history_rnn <- model_rnn$fit(
  x = x_train2,
  y = y_train2,
  epochs = r_to_py(100L),
  batch_size = r_to_py(16L),
  verbose = 0
)
```

```{r}
# Prepare input for forecasting
input_seq <- model_data[(length(model_data)-10):(length(model_data)-1)] # 10 points before test

input_seq_array <- array(as.numeric(input_seq), dim = c(1, timesteps,1))

input_seq_array <- tf$convert_to_tensor(input_seq_array, dtype = "float32")

# Prediction function
predict_next <- function(model, input_seq_array, n_steps = 10) {
  preds <- c()
  current_input <- input_seq_array
  for (i in 1:n_steps) {
    pred <- model$predict(current_input)
    preds <- c(preds, as.numeric(pred))
    updated_seq <- c(as.numeric(current_input)[-1], as.numeric(pred))
    current_input <- array(updated_seq, dim = c(1, timesteps, 1))
    current_input <- tf$convert_to_tensor(current_input, dtype = "float32")
  }
  return(preds)
}
```

```{r}
lstm_preds_norm <- predict_next(
  model_lstm, 
  input_seq_array, 
  n_steps = 10
)
```

```{r}
rnn_preds_norm <- predict_next(
  model_rnn, 
  input_seq_array, 
  n_steps = 10
)
```

Scale Back

```{r}
# Initialize vectors to store scaled-back predictions and true prices
lstm_preds_original <- numeric(10)
true_prices_original <- numeric(10)
rnn_preds_original <- numeric(10)

# Scale back one by one (because each day uses its own lagged High/Low)
for (i in 1:10) {
  lagged_min <- data2$AMZN.Low[length(train_data2) + i - 1] # lag(Low) for that day
  lagged_max <- data2$AMZN.High[length(train_data2) + i - 1] # lag(High) for that day
  rnn_preds_original[i] <- rnn_preds_norm[i] * (lagged_max - lagged_min) + lagged_min
  lstm_preds_original[i] <- lstm_preds_norm[i] * (lagged_max - lagged_min) + lagged_min
  true_prices_original[i] <- data2$AMZN.Close[length(train_data2) + i] # true Close at day
}
```

MSE

```{r}
mse_rnn <- mean((rnn_preds_original - true_prices_original)^2)

mse_lstm <- mean((lstm_preds_original - true_prices_original)^2)

cat("Test MSE (RNN, original scale):", mse_rnn, "\n")
cat("Test MSE (LSTM, original scale):", mse_lstm, "\n")
```

Plot 10-Day Forecast (Original Scale)

```{r}
plot(
  (length(train_data2)+1):(length(train_data2)+10),
  true_prices_original,
  type = 'l', 
  lwd = 2, 
  col = 'black', 
  xlab = "Time Index", 
  ylab = "Close Price (USD)",
  main = "10-Day Forecast vs True (Original Scale)"
)

lines(
  (length(train_data2)+1):(length(train_data2)+10),
  rnn_preds_original, 
  col = 'red', 
  lwd = 2
)

lines(
  (length(train_data2)+1):(length(train_data2)+10),
  lstm_preds_original, 
  col = 'blue', 
  lwd = 2
)

legend(
  "bottomleft", 
  legend = c("True Close", "RNN Prediction", "LSTM Prediction"),
  col = c("black", "red", "blue"), 
  lwd = 2
)
```

## Comparison

Model Test MSE (Original USD Scale)

-   RNN: 16.05965

-   LSTM: 14.42587

-   LSTM has lower MSE than RNN. LSTM outperforms RNN when predicting actual stock prices.

The reason why LSTM outperform RNN

-   The algorithm difference between LSTM and RNN. LSTM is known to be improvement over RNN because LSTM has separate memory cells and three gates.

-   Vanish/Exploding Gradient Problem, one biggest issue that RNN has

    -   When we train RNN, Gradient can vanish or it can explode, either one would result in bad prediction
