---
title: "AMS 580 Quiz 2"
author: "Kachun Huang"
date: "2025-02-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# AMS 580 Quiz Logistic Regression

## Library

```{r}
library(MASS)
library(tidyverse)
library(leaps)
library(car)
library(caret)
library(dplyr)
library(lmtest)
```

## Data import

```{r}
titantic <- read.table("/Users/thomashuang/Downloads/Titanic2.csv", sep = ",",header=TRUE)
head(titantic)
```

```{r}
str(titantic)
```

## Data pre-processing

```{r}
#exlcude particular columns
titantic <- subset(titantic, select = -c(PassengerId,Name,Ticket,Cabin))
str(titantic)
```

```{r}
#omit missing values
titantic <- na.omit(titantic)
str(titantic)
```

```{r}
#factorize survivied and Pclass column
titantic$Survived <- factor(titantic$Survived)
titantic$Pclass <- factor(titantic$Pclass)
str(titantic)
```

After cleaning, there are only 714 observations left.

## Data Splitting

```{r}
set.seed(123)
training.samples <- titantic$Survived %>% createDataPartition(p=0.8, list=FALSE)
train.data <- titantic[training.samples,]
test.data <- titantic[-training.samples,]
```

## Model fitting

```{r}
#fit the model
fit <- glm(Survived ~., data=train.data, family = binomial)

#summary
summary(fit)
```

## Confusion Matrix

#### Manual Calculation

```{r}
# find Y hat of each x
pred_prob <- fit %>% predict(test.data, type="response")

#create predicted class
predicted.classes <- ifelse(pred_prob > 0.5, "survived", "dead")

test.data$Survived <- ifelse(test.data$Survived == 1, "survived", "dead")

#confusion matrix 
table(predicted.classes,test.data$Survived)
```

#### Built-in Function

```{r}
cm <- confusionMatrix(factor(predicted.classes),factor(test.data$Survived), positive = "survived")
```

When positive class = "1" Detect survived

\| Actual/Predicted \| Negative (Actual 0) \| Positive (Acutal 1) \|

\| Negative (Predicted 0) \| **True Negative (TN)** \| **False Negative(FN)** \|

\| Positive (Predicted 1) \| **False Positive(FP)** \| **True Positive(TP)** \|

### Overall accuracy

$P(\text{Predicted Correctly}) = \frac{TP+TN}{TP+TN+FP+FN}$

```{r}
(71+43)/(71+15+13+43)
```

```{r}
cm$overall["Accuracy"]
```

### Sensitivity

$Sensitivity= P(\text{true positive}|\text{correct}) = \frac{TP}{TP+FN}$

```{r}
(43)/(43+15)
```

```{r}
cm$byClass["Sensitivity"]
```

### Specificity

$Specificity= P(\text{true negative}|\text{correct})=\frac{TN}{TN+FP}$

```{r}
(71)/(13+71)
```

```{r}
cm$byClass["Specificity"]
```

## Prediction

```{r}
new_data <- data.frame(
  Pclass = factor(c(3,1,2)),
  Sex = c("male","female","male"),
  Age = c(24,68,41),
  SibSp = c(1,0,1),
  Parch = c(0,0,2),
  Fare = c(8.42,24.34,41.93),
  Embarked=c("Q","C","S")
)
new_data
```

```{r}
# find Y hat of a given X
pred_fit <- predict(fit, new_data, type="response")
pred_fit
result <- ifelse(pred_fit > 0.5, "survived","dead")
result
```

-   Passenger ID 892 will not survived.

-   Passenger ID 893 will survived.

-   Passenger ID 894 will no survived
