---
title: "Penalized Regression"
author: "Kachun Huang"
date: "2025-04-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Penalized Regression in R

## Import libraries

```{r}
library(glmnet)
library(caret)
library(MASS)
library(tidyverse)
```

## Model Matrix

```{r eval=FALSE, include=TRUE}
x <- model.matrix(response ~., train.data)[,-1]
y <- train.data$response
```

## Ridge Regression

Ridge regression adds a penalty equal to the square of the magnitude of coefficients, which shrinks the coefficients towards zero but does not set them exactly to zero:

```{r eval=FALSE, include=TRUE}
# Find the best lambda using cross-validation
cv.ridge <- cv.glmnet(x, y, alpha = 0)
```

### Best Lambda 

```{r eval=FALSE, include=TRUE}
best_lambda <- cv.ridge$lambda.min
```

### Fit the Model with best lambda

```{r eval=FALSE, include=TRUE}
# Fit the model with the best lambda
ridge_model <- glmnet(x, y, alpha = 0, lambda = best_lambda)
```

#### Display coefficients

```{r eval=FALSE, include=TRUE}
coef(ridge_model)
```

## Lasso Regression

Lasso regression uses the L1 penalty, which can shrink some coefficients to exactly zero, effectively performing variable selection:

```{r eval=FALSE, include=TRUE}
# Find the best lambda using cross-validation
cv.lasso <- cv.glmnet(x, y, alpha = 1)
```

### Best Lambda 

```{r eval=FALSE, include=TRUE}
best_lambda <- cv.lasso$lambda.min
```

### Fit the Model with best lambda

```{r eval=FALSE, include=TRUE}
# Fit the model with the best lambda
lasso_model <- glmnet(x, y, alpha = 1, lambda = best_lambda)
```

#### Display coefficients

```{r eval=FALSE, include=TRUE}
coef(lasso_model)
```

## Elastic Net Regression

Elastic Net combines the properties of both Ridge and Lasso by using both L1 and L2 penalties:

We need to find the optimal $\alpha$ and $\lambda$

```{r eval=FALSE, include=TRUE}
model <- train(
  repsonse~.,
  data = train.data,
  method = "glmnet",
  trControl = trainControl("cv", number=10),
  tuneLength = 10
)
```

### Best Alpha and Lambda 

```{r eval=FALSE, include=TRUE}
model$bestTune
```

### Fit the Model with best lambda

```{r eval=FALSE, include=TRUE}
# Fit the model with the best lambda and alpha
elastic_model <- glmnet(x, y, alpha = best_alpha, lambda = best_lambda)
```

#### Display coefficients

```{r eval=FALSE, include=TRUE}
coef(elastic_model)
```

# Example

## Import Data

```{r}
data("Boston", package="MASS")
str(Boston)
```

## Data Splitting

```{r}
set.seed(123)
training.samples <- Boston$medv %>% createDataPartition(p=0.75, list=FALSE)
train.data <- Boston[training.samples,]
test.data <- Boston[-training.samples,]
```

## Model Matrix

```{r}
x <- model.matrix(medv~., train.data)[,-1] #exclude the last column (medv)
y <- train.data$medv
```

## Fit Ridge Regression

```{r}
cv <- cv.glmnet(
 x,
 y,
 alpha=0
)
```

### Best Lambda 

```{r}
cv$lambda.min
```

### Fit the Model with best lambda

```{r}
model <- glmnet(
  x,
  y,
  alpha=0,
  lambda = cv$lambda.min
)
```

#### Model Coefficient

```{r}
coef(model)
```

### Prediction

```{r}
x.test <- model.matrix(medv~., test.data)[,-1]
pred_ridge <- model %>% predict(x.test) %>% as.vector()
```

### RMSE and R\^2

```{r}
data.frame(
  RMSE = RMSE(pred_ridge, test.data$medv),
  Rsquare = R2(pred_ridge, test.data$medv)
)
```

## Fit Lasso Regression

```{r}
cv <- cv.glmnet(
 x,
 y,
 alpha=1
)
```

### Best Lambda 

```{r}
cv$lambda.min
```

### Fit the Model with best lambda

```{r}
model <- glmnet(
  x,
  y,
  alpha=1,
  lambda = cv$lambda.min
)
```

#### Model Coefficient

```{r}
coef(model)
```

### Prediction

```{r}
x.test <- model.matrix(medv~., test.data)[,-1]
pred_lasso <- model %>% predict(x.test) %>% as.vector()
```

### RMSE and R\^2

```{r}
data.frame(
  RMSE = RMSE(pred_lasso, test.data$medv),
  Rsquare = R2(pred_lasso, test.data$medv)
)
```

## Fit Elastic Net Regression

```{r}
model <- train(
  medv~.,
  data=train.data,
  method="glmnet",
  trControl=trainControl("cv",number=10),
  tuneLength=10
)
```

### Best Alpha and Lambda 

```{r}
model$bestTune
```

### Fit the Model with best lambda and alpha

```{r}
model <- glmnet(
  x,
  y,
  alpha=model$bestTune$alpha,
  lambda=model$bestTune$lambda
)
```

#### Model Coefficient

```{r}
coef(model)
```

### Prediction

```{r}
x.test <- model.matrix(medv~., test.data)[,-1]
pred_en <- model %>% predict(x.test) %>% as.vector()
```

### RMSE and R\^2

```{r}
data.frame(
  RMSE = RMSE(pred_en, test.data$medv),
  Rsquare = R2(pred_en, test.data$medv)
)
```

## Ridge Regression using train()

```{r}
set.seed(123)

grid_ridge <- expand.grid(
  alpha=0,
  lambda = seq(0.001,0.1, length=100)
)

model_ridge <- train(
  medv~.,
  data=train.data,
  method="glmnet",
  metric="RMSE",
  trConrol=trainControl(method="cv",number=10),
  tuneGrid=grid_ridge
)
```

### Best Tune

```{r}
model_ridge$bestTune
```

#### Model Coefficient

```{r}
coef(model_ridge)
```

### Prediction

```{r}
x.test <- model.matrix(medv~., test.data)[,-1]
pred_ridge <- model_ridge %>% predict(x.test) %>% as.vector()
```

### RMSE and R\^2

```{r}
data.frame(
  RMSE = RMSE(pred_ridge, test.data$medv),
  Rsquare = R2(pred_ridge, test.data$medv)
)
```

## 

## Lasso Regression using train()

```{r}
set.seed(123)

grid_lasso <- expand.grid(
  alpha=1,
  lambda = seq(0.001,0.1, length=100)
)

model_lasso <- train(
  medv~.,
  data=train.data,
  method="glmnet",
  metric="RMSE",
  trConrol=trainControl(method="cv",number=10),
  tuneGrid=grid_lasso
)
```

### Best Tune

```{r}
model_lasso$bestTune
```

#### Model Coefficient

```{r}
coef(model_lasso)
```

### Prediction

```{r}
x.test <- model.matrix(medv~., test.data)[,-1]
pred_lasso <- model_lasso %>% predict(x.test) %>% as.vector()
```

### RMSE and R\^2

```{r}
data.frame(
  RMSE = RMSE(pred_lasso, test.data$medv),
  Rsquare = R2(pred_lasso, test.data$medv)
)
```

## 
