---
title: "AMS 580 Quiz 4"
author: "Kachun Huang"
date: "2025-04-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# AMS 580 Quiz 4

## Import Library

```{r}
library(dplyr)
library(caret)
library(randomForest)
library(tidyverse)
library(rpart)
library(rpart.plot)
library(rattle)
library(caTools)
```

# Regression Task

## Import Data

```{r}
house <- read.csv("/Users/thomashuang/Downloads/Ames_Housing_Data.csv")
head(house)
```

```{r}
str(house)
```

## Remove Missing Values

```{r}
house <- na.omit(house)
str(house)
```

After removing the missing value, there are 1460 obs left.

## Data Splitting

```{r}
set.seed(123)
training.samples <- house$SalePrice %>% createDataPartition(p=0.75, list=FALSE)
train.data <- house[training.samples,]
test.data <- house[-training.samples,]
```

## Prune CART

```{r}
set.seed(123)

model_CART <- train(
  SalePrice ~., 
  data = train.data, 
  method = "rpart",
  trControl = trainControl("cv", number = 10)
)
```

### Access the best tune

```{r}
model_CART$bestTune
```

The best CP value is 0.07317

### Plot the best model

```{r}
fancyRpartPlot(model_CART$finalModel)
```

### Prediction

```{r}
pred_CART <- predict(
  model_CART,
  test.data
)
```

### Test RMSE

```{r}
RMSE(pred_CART,test.data$SalePrice)
```

## Best Random Forest

```{r}
set.seed(123)

# 10-fold CV
control <- trainControl(
  method = "cv", 
  number = 10
  )

# train the rf using 10-fold CV
model_rf <- train(
  SalePrice~., 
  data = train.data, 
  method = "rf", 
  trControl = control
  )

print(model_rf)
```

### Access bestTune

```{r}
print(model_rf$bestTune)
```

The best mtry value is 10

### Plot importance measures

```{r}
varImpPlot(model_rf$finalModel)
```

### Prediction

```{r}
pred_rf <- predict(
  model_rf,
  newdata = test.data
)
```

### Test RMSE

```{r}
RMSE(pred_rf, test.data$SalePrice)
```

## Which is better?

The Test RMSE of CART is 50958.41

The Test RMSE of Random Forest is 37341

In terms of test RMSE, random forest performs better than CART.

The possible reasons of random forest performs better than CART are

-   CART generally overfit on training data, result in high rmse when evaluate with test data

-   Random Forest performs better because of its algorithm

    -   by aggregating all the prediction of all tree created from bootstrap sample

    -   It allows the model to foresee unseen data, so lower the rmse when evaluate with test data

# Classification Task

## Import Data

```{r}
titanic <- read.csv("/Users/thomashuang/Downloads/Titanic.csv")
```

```{r}
head(titanic)
```

```{r}
str(titanic)
```

## Exclude some columns

```{r}
titanic <- subset(titanic, select = -c(PassengerId, Name, Ticket, Cabin))
```

```{r}
str(titanic)
```

## Remove Missing Values

```{r}
titanic <- na.omit(titanic)
str(titanic)
```

## Factorize Column

```{r}
titanic$Survived <- as.factor(titanic$Survived)
titanic$Pclass <- as.factor(titanic$Pclass)
```

```{r}
str(titanic)
```

## Data Splitting

```{r}
set.seed(123)
training.samples <- titanic$Survived %>% createDataPartition(p=0.75, list=FALSE)
train.data <- titanic[training.samples,]
test.data <- titanic[-training.samples,]
```

## Prune CART

```{r}
set.seed(123)

model_CART_c <- train(
  Survived ~., 
  data = train.data, 
  method = "rpart",
  trControl = trainControl("cv", number = 10)
)
```

### Access the best tune

```{r}
model_CART_c$bestTune
```

The best CP value is 0.01376147

### Plot the best model

```{r}
fancyRpartPlot(model_CART_c$finalModel)
```

### Prediction

```{r}
pred_CART_c <- predict(
  model_CART_c,
  test.data
)
```

### Confusion Matrix

```{r}
cm1 <- confusionMatrix(
  factor(pred_CART_c),
  factor(test.data$Survived),
  positive = "1"
)
cm1
```

#### Sensitivity

```{r}
cm1$byClass["Sensitivity"]
```

#### Specificity

```{r}
cm1$byClass["Specificity"]
```

#### Overall Accuracy

```{r}
cm1$overall["Accuracy"]
```

## Best Random Forest

```{r}
set.seed(123)
# 10-fold CV
control <- trainControl(
  method = "cv", 
  number = 10
  )

# train the rf using 10-fold CV
model_rf_c <- train(
  Survived ~ ., 
  data = train.data, 
  method = "rf", 
  trControl = control, 
  importance = TRUE
  )

print(model_rf_c)
```

### Access bestTune

```{r}
model_rf_c$bestTune
```

The best mtry is 2

### Plot importance measures

```{r}
varImpPlot(model_rf_c$finalModel)
```

### Prediction

```{r}
pred_rf_c <- predict(
  model_rf_c,
  newdata = test.data
)
```

### Confusion Matrix

```{r}
cm2 <- confusionMatrix(
  pred_rf_c,
  test.data$Survived,
  positive = "1"
)
cm2
```

#### Sensitivity

```{r}
cm2$byClass["Sensitivity"]
```

#### Specificity

```{r}
cm2$byClass["Specificity"]
```

#### Overall Accuracy

```{r}
cm2$overall["Accuracy"]
```

## Which is better?

In terms of sensitivity, Random Forest is better

In terms of specificity, CART is beter

In terms of overall accuracy, CART is better

The reason behind the difference in performance:

-   Generally Random Forest relies on sub-seting random features to create multiple CART model. This dataset only has 7 features, so the Random Forest doesn't get much exposure to more diverse data.
