---
title: "AMS 580 Logistic Regression HW"
author: "Kachun Huang"
date: "2025-02-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# AMS 580 Logistic Regression HW

Logistic Regression (\* A type of Generalized Linear Model) with the Banknote Data The\
banknote.csv data (see attached) were extracted from images that were taken from\
genuine and forged banknote-like specimens. Yes, this is a Catch Me if You Can story. For\
digitization, an industrial camera usually used for print inspection was used. The final\
images have 400x 400 pixels. Wavelet Transform tool were used to extract features from\
images. There are 1372 banknotes, and 5 variables:\
1. variance of Wavelet Transformed image (continuous)\
2. skewness of Wavelet Transformed image (continuous)\
3. curtosis of Wavelet Transformed image (continuous)\
4. entropy of image (continuous)\
5. class (binary) – this is the response variable of interest, 1 (forged) or 0 (genuine).\

## Import library

```{r}
library(tidyverse)
library(caret)
library(MASS)
```

## Data Import

```{r}
banknote <- read.csv("/Users/thomashuang/Downloads/banknote.csv")
str(banknote)
```

## Data pre-processing

### Handle Missing Value

```{r}
banknote <- na.omit(banknote)
```

### Factor

Since class should be a `factor` type not `int` type, we need to convert from `int` to `factor` .

```{r}
banknote$class <- factor(banknote$class, levels = c(0,1))
str(banknote)
```

## Data Splitting

Please split the data into 80% training and 20% testing using seed =123.

```{r}
set.seed (123)
training.samples <- banknote$class %>% createDataPartition(p=0.8, list=FALSE)
train.data <- banknote[training.samples,]
test.data <- banknote[-training.samples,]
```

## Model Fitting

Then you shall fit a logistic regression model with all 4 predictors using the training data.

```{r}
full.model <- glm(class ~., family = binomial, data = train.data)

summary(full.model)
```

## Prediction

(c) Please use this fitted model based on the training data to predict the response variable ‘class’ (whether the banknote is forged or not) for the testing data.

```{r}
pred_prob <- full.model %>% predict (test.data, type="response")
predicted.classes <- ifelse(pred_prob > 0.5, 1, 0)
```

## Confusion Matrix

Please generate the confusion matrix, and report: (i) The overall accuracy;

```{r}
confu_matrix <- confusionMatrix(factor(predicted.classes), factor(test.data$class), positive ="1")
confu_matrix
```

When positive class = "1" Detect forged

\| Actual/Predicted \| Negative (Actual 0) \| Positive (Acutal 1) \|

\| Negative (Predicted 0) \| **True Negative (TN)** \| **False Negative(FN)** \|

\| Positive (Predicted 1) \| **False Positive(FP)** \| **True Positive(TP)** \|

## Sensitivity

(ii) The sensitivity (that is, the probability a banknote is predicted to be forged given that it was in fact forged);

```{r}
confu_matrix$byClass["Sensitivity"]
```

## Specificity

(iii) The specificity (that is, the probability a banknote is predicted to be genuine given that it was in fact genuine).

```{r}
confu_matrix$byClass["Specificity"]
```

## Accuracy

```{r}
confu_matrix$overall["Accuracy"]
```
