---
title: "AMS 580 Quiz 1"
author: "Kachun Huang"
date: "2025-02-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# AMS 580 Quiz 1

```{r}
library(MASS)
library(tidyverse)
library(leaps)
library(car)
library(caret)
library(dplyr)
```

## Data Import

```{r}
house <- read.table("/Users/thomashuang/Downloads/Ames_Housing_Data.csv", sep=",",header=TRUE)
```

```{r}
head(house)
```

```{r}
#remove Id column
house <- subset(house, select = -c(Id))

head(house)
```

## Data Split

```{r}
set.seed(123)
training.samples <- house$SalePrice %>% createDataPartition(p=0.75,list = FALSE)
train.data <- house[training.samples, ]
test.data <- house[-training.samples, ]
```

## Step-wise BIC

```{r}
# null model
null <- lm(SalePrice ~ 1, data=train.data)

#full model
full <-lm(SalePrice ~. , data=train.data)
```

### Forward

```{r}
#find the best model in forward stepwise
fit_step_fwd_bic <- stepAIC(null, scope = list (lower = null, upper = full),
direction = "forward", k = log(nrow(train.data)), trace = 1)
```

```{r}
#display the BIC
BIC(fit_step_fwd_bic)
```

### Backward

```{r}
#find the best model in backward stepwise
fit_step_back_bic <- stepAIC(full, k = log(nrow(train.data)), trace = 1,direction = "backward")
fit_step_back_bic
```

```{r}
#display the BIC 
BIC(fit_step_back_bic)
```

### Both

```{r}
#find the best model in backward stepwise
fit_step_both_bic <- stepAIC(full, k = log(nrow(train.data)), trace = 1,direction="both")
fit_step_both_bic
```

```{r}
#display the BIC 
BIC(fit_step_both_bic)
```

### Step-wise Result

Since Backward yield the smallest BIC.

### Coefficients

```{r}
summary(fit_step_back_bic)$coefficients[,1]
```

### RMSE and R\^2

```{r}
pred_step_back_bic <- fit_step_back_bic %>% predict(test.data)
data_frame(
  RMSE = RMSE(pred_step_back_bic, test.data$SalePrice),
  Rsquare = R2(pred_step_back_bic, test.data$SalePrice)
)
```

## Best Subset SSE

```{r}
# use best subset to find the model
fit_bs <- regsubsets(SalePrice ~., data=train.data, nvmax = 30)

#save summary
result <- summary(fit_bs)
```

```{r}
#display the sse of each model candidate
result$rss
```

```{r}
#find which model that have the least sse
which.min(result$rss)
```

```{r}
#sse of the 18th model
result$rss[18]
```

From this, we know the best model based on the SSE criterion is the model with 18 predictors.

```{r}
result$which[18,]
```

The result above indicate that the full model has the lowest SSE of 1.16219e+12.

### Coefficients

```{r}
#fit the full model
fit_full <- lm(SalePrice ~. , data=train.data)

#the coefficient of full model
summary(fit_full)$coefficients[,1]
```

### RMSE and R\^2

```{r}
pred_full <- fit_full %>% predict(test.data)
data_frame(
  RMSE = RMSE(pred_full, test.data$SalePrice),
  Rsquare = R2(pred_full, test.data$SalePrice)
)
```

## Which is the best?

Which model selection method among the 2 we have used above is the best?

Step-wise selection method performs the best.

(a) Please compare the BIC of these models using the training data, as well as display these two models so we can see the parameter estimators and model goodness of fit measures.

```{r}
#Stepwise best BIC
BIC(fit_step_back_bic)
```

```{r}
#Stepwise best summary
summary(fit_step_back_bic)
```

```{r}
#best subset best BIC
BIC(full)
```

```{r}
#best subset best
summary(full)
```

(b) Furthermore, please compare the RMSE and R2 of these models using the test data.

```{r}
# RMSE and R2 of step-wise
pred_step_back_bic <- fit_step_back_bic %>% predict(test.data)
data_frame(
  RMSE = RMSE(pred_step_back_bic, test.data$SalePrice),
  Rsquare = R2(pred_step_back_bic, test.data$SalePrice)
)
```

```{r}
# RMSE and R2 of best subset
pred_full <- fit_full %>% predict(test.data)
data_frame(
  RMSE = RMSE(pred_full, test.data$SalePrice),
  Rsquare = R2(pred_full, test.data$SalePrice)
)
```

(c) Please discuss any modifications you can do to further improve your model(s).

For improvement, I won't use SSE criterion for model selection since full model generally has the least SSE which doesn't account for model complexity.
